---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# StanEstimators

<!-- badges: start -->
[![R-CMD-check](https://github.com/andrjohns/StanEstimators/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/andrjohns/StanEstimators/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

The `StanEstimators` package provides an estimation back-end for R functions,
similar to those provided by the `optim` package, using the algorithms provided
by the Stan probabilistic programming language.

As Stan's algorithms are gradient-based, function gradients can be automatically
calculated using finite-differencing or the user can provide a function for
analytical calculation.

## Usage

Consider the goal of estimating the mean and standard deviation of a normal
distribution, with uniform uninformative priors on both parameters:

$$
y \sim \textbf{N}(\mu, \sigma)
$$

$$
\mu \sim \textbf{U}[-\infty, \infty]
$$

$$
\sigma \sim \textbf{U}[0, \infty]
$$

With known true values for verification:

```{r}
y <- rnorm(500, 10, 2)
```


As with other estimation routines provided in R, we need to specify this as a
function which takes a vector of parameters as its first argument and returns a
single scalar value (the log-likelihood), as well as initial values for the
parameters:

```{r}
loglik_fun <- function(v, x) {
  sum(dnorm(x, v[1], v[2], log = TRUE))
}

inits <- c(0, 5)
```

Estimation time can also be significantly reduced by providing a gradient function,
rather than relying on finite-differencing:

```{r}
grad <- function(v, x) {
  inv_sigma <- 1 / v[2]
  y_scaled = (x - v[1]) * inv_sigma
  scaled_diff = inv_sigma * y_scaled
  c(sum(scaled_diff),
    sum(inv_sigma * (y_scaled*y_scaled) - inv_sigma)
  )
}
```

### MCMC Estimation

Full MCMC estimation is provided by the `stan_sample()` function, which uses
Stan's default No U-Turn Sampler (NUTS) unless otherwise specified:

```{r, results=FALSE, message=FALSE, warning=FALSE}
library(StanEstimators)

fit <- stan_sample(loglik_fun, inits, y,
                   lower = c(-Inf, 0), # Enforce a positivity constraint for SD
                   num_chains = 1,
                        seed = 1234)
```

We can see that the parameters were recovered accurately and that the estimation
was relatively fast: ~1 sec for 1000 warmup and 1000 iterations

```{r}
unlist(fit@timing)
summary(fit)
```

Estimation time can be improved further by providing a gradient function:
```{r, results=FALSE, message=FALSE, warning=FALSE}
fit_grad <- stan_sample(loglik_fun, inits, y,
                        grad_fun = grad,
                        lower = c(-Inf, 0),
                        num_chains = 1,
                        seed = 1234)
```

Which shows that the estimation time was dramatically improved,
now ~0.15 seconds for 1000 warmup and 1000 iterations.

```{r}
unlist(fit_grad@timing)
summary(fit_grad)
```

### Optimization

```{r, results=FALSE, message=FALSE, warning=FALSE}
opt_fd <- stan_optimize(loglik_fun, inits, y,
                          lower = c(-Inf, 0),
                          seed = 1234)
opt_grad <- stan_optimize(loglik_fun, inits, y,
                          grad_fun = grad,
                          lower = c(-Inf, 0),
                          seed = 1234)
```

```{r}
summary(opt_fd)
summary(opt_grad)
```

### Laplace Approximation

```{r, results=FALSE, message=FALSE, warning=FALSE}
# Can provide the mode as a numeric vector: 
lapl_num <- stan_laplace(loglik_fun, inits, y,
                          mode = c(10, 2),
                          lower = c(-Inf, 0),
                          seed = 1234)

# Can provide the mode as a StanOptimize object: 
lapl_opt <- stan_laplace(loglik_fun, inits, y,
                          mode = opt_fd,
                          lower = c(-Inf, 0),
                          seed = 1234)

# Can estimate the mode before sampling: 
lapl_est <- stan_laplace(loglik_fun, inits, y,
                          lower = c(-Inf, 0),
                          seed = 1234)
```

```{r}
summary(lapl_num)
summary(lapl_opt)
summary(lapl_est)
```

### Variational Inference

```{r, results=FALSE, message=FALSE, warning=FALSE}
var_fd <- stan_variational(loglik_fun, inits, y,
                              lower = c(-Inf, 0),
                              seed = 1234)
var_grad <- stan_variational(loglik_fun, inits, y,
                              grad_fun = grad,
                              lower = c(-Inf, 0),
                              seed = 1234)
```

```{r}
summary(var_fd)
summary(var_grad)
```

### Pathfinder

```{r, results=FALSE, message=FALSE, warning=FALSE}
path_fd <- stan_pathfinder(loglik_fun, inits, y,
                              lower = c(-Inf, 0),
                              seed = 1234)
path_grad <- stan_pathfinder(loglik_fun, inits, y,
                              grad_fun = grad,
                              lower = c(-Inf, 0),
                              seed = 1234)
```

```{r}
summary(path_fd)
summary(path_grad)
```




