---
title: "Getting Started with StanEstimators"
output:
  html_vignette:
    toc: yes
author: "Andrew Johnson"
date: "`r Sys.Date()`"
vignette: >
  %\VignetteIndexEntry{Getting Started with StanEstimators}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(StanEstimators)
```

# Introduction

This vignette provides a basic introduction to the features and capabilities of the `StanEstimators` package. `StanEstimators` provides an interface for estimating R functions using the various algorithms and methods implemented by [Stan](https://mc-stan.org/).

The `StanEstimators` package supports all algorithms implemented by Stan. The available methods, and their corresponding functions, are:

Stan Method | `StanEstimators` Function
----------- | -------------------------
MCMC Sampling | `stan_sample`
Maximum Likelihood Estimation | `stan_optimize`
Variational Inference | `stan_variational`
Pathfinder | `stan_pathfinder`
Laplace Approximation | `stan_laplace`

## Motivations

While Stan is powerful, it can have a high barrier to entry for new users - as they need to translate their existing models/functions into the Stan language. `StanEstimators` provides a simple interface for users to estimate their R functions using Stan's algorithms without needing to learn the Stan language. Additionally, it also allows for users to 'sanity-check' their Stan code by comparing the results of their Stan code to the results of their original R function. Finally, it can be difficult to interface Stan with existing R packages and functions - as this requires bespoke Stan models for the problem at hand, something that may be too great a time investment for many users.

`StanEstimators` aims to address these issues.

## Estimating a Model

As an example, we will use the popular 'Eight Schools' example from the Stan documentation. This example is used to demonstrate the use of hierarchical models in Stan. The model is defined as:

$$
y_j &\sim N(\theta_j, \sigma_j), \quad j=1,\ldots,8 \\
\theta_j &\sim \N(\mu, \tau), \quad j=1,\ldots,8
$$

With data:

School | Estimate ($y_j$) | Standard Error ($\sigma_j$)
------ | -------- | --------------
A      | 28       | 15
B      | 8        | 10
C      | -3       | 16
D      | 7        | 11
E      | -1       | 9
F      | 1        | 11
G      | 18       | 10
H      | 12       | 18

```{r}
y <- c(28,  8, -3,  7, -1,  1, 18, 12)
sigma <- c(15, 10, 16, 11,  9, 11, 10, 18)
```

### Specifying the Function

To specify this as a function compatible with `StanEstimators`, we need to define a function that takes in a vector of parameters as the first argument and returns a single value (generally the joint log-likelihood):

```{r}
eight_schools_lpdf <- function(v, y, sigma) {
  mu <- v[1]
  tau <- v[2]
  eta <- v[3:10]

  # Use the non-centered parameterisation for eta
  # https://mc-stan.org/docs/stan-users-guide/reparameterization.html
  theta <- mu + tau * eta

  sum(dnorm(eta, mean = 0, sd = 1, log = TRUE)) +
    sum(dnorm(y, mean = theta, sd = sigma, log = TRUE))
}
```

Note that any additional data required by the function are passed as additional arguments. In this case, we need to pass the data for $y$ and $\sigma$. Alternatively, the function can assume that these data will be available in the global environment, rather than passed as arguments.

### Estimating the Function

To estimate our model, we simply pass the function to the relevant `StanEstimators` function. For example, to estimate the function using MCMC sampling, we use the `stan_sample` function (which uses the No-U-Turn Sampler by default).

#### Parameter Bounds

Because we are estimating a standard deviation in our model ($\tau$), we need to ensure that it is positive. We can do this by specifying a lower bound of 0 for $\tau$. This is done by passing a vector of lower bounds to the `lower` argument, with the corresponding elements of the vector matching the order of the parameters in the function. Noting that $\tau$ is the second parameter in the function, and we do not want to specify a lower bound for any other parameters, we can specify the lower bounds as:

```{r}
lower <- c(-Inf, 0, rep(-Inf, 8))
```


#### Running the Model

We can now pass these arguments to the `stan_sample` function to estimate our model. We will use the default number of warmup iterations (1000) and sampling iterations (1000).

Note that we need to specify the number of parameters in the model (10) using the `n_pars` argument. This is because the function does not know how many parameters are in the model, and so cannot automatically determine this.

```{r, results=FALSE, message=FALSE, warning=FALSE}
fit <- stan_sample(eight_schools_lpdf,
                   n_pars = 10,
                   additional_args = list(y = y, sigma = sigma),
                   lower = lower,
                   num_chains = 1)
```

### Inspecting the Results

The estimates are stored in the `fit` object using the [`posterior::draws` format](https://CRAN.R-project.org/package=posterior/vignettes/posterior.html). We can inspect the estimates using the `summary` function (which calls `posterior::summarise_draws`):

```{r}
summary(fit)
```

## Model Checking and Comparison - Leave-One-Out Cross-Validation (LOO-CV)

`StanEstimators` also supports the use of the [loo](https://mc-stan.org/loo/articles/loo2-example.html) package for model checking and comparison. To use this, we need to specify a function which returns the pointwise log-likelihood for each observation in the data - as our original function returns the sum of all log-likelihoods.

For our model, we can define this function as:

```{r}
eight_schools_pointwise <- function(v, y, sigma) {
  mu <- v[1]
  tau <- v[2]
  eta <- v[3:10]

  # Use the non-centered parameterisation for eta
  # https://mc-stan.org/docs/stan-users-guide/reparameterization.html
  theta <- mu + tau * eta

  # Only the log-likelihood for the outcome variable
  dnorm(y, mean = theta, sd = sigma, log = TRUE)
}
```

This can then be used with the `loo` function to calculate the LOO-CV estimate:

```{r}
loo(fit, pointwise_ll_fun = eight_schools_pointwise,
    additional_args = list(y = y, sigma = sigma))
```
