---
title: "Automatic Differentiation with RTMB"
output:
  html_vignette:
    toc: yes
    toc_depth: 3
author: "Andrew Johnson"
date: "`r Sys.Date()`"
vignette: >
  %\VignetteIndexEntry{Automatic Differentiation with RTMB}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

```{r setup}
library(StanEstimators)
```

# Introduction

Stan's algorithms, including MCMC sampling (NUTS), optimization (L-BFGS), variational inference, Pathfinder, and Laplace approximation, are gradient-based methods. This means they require not only the log-probability function but also its gradient (the vector of partial derivatives with respect to each parameter) to work efficiently.

`StanEstimators` provides three ways to compute gradients:

1. **Finite differences** (default): Automatic but slow. Approximates gradients by evaluating the function at slightly perturbed parameter values.
2. **Analytical gradients**: Fast and accurate, but requires you to manually derive and code the gradient function.
3. **RTMB automatic differentiation**: Fast and automatic. Uses the `RTMB` package to compute exact gradients via automatic differentiation (AD).


## Installing RTMB

To use RTMB with `StanEstimators`, you need to install the `RTMB`, `withr`, and `future` packages:

```{r, eval=FALSE}
install.packages(c("RTMB", "withr", "future"))
```

Once installed, simply set `grad_fun = "RTMB"` in any `StanEstimators` function to enable automatic differentiation.

For basic usage of `StanEstimators`, see the [Getting Started](Getting-Started.html) vignette.

# Poisson Regression

Next, we'll examine a generalized linear model (GLM) for count data. Poisson regression uses a log-link function: $\log(\lambda_i) = X_i\beta$, where $\lambda_i$ is the expected count.

## Simulating Data

```{r}
set.seed(456)
n <- 200
p <- 2  # number of predictors (plus intercept)

# True coefficients
true_beta <- c(0.5, 1.2, -0.8)

# Design matrix (intercept + 2 predictors)
X <- cbind(1, matrix(rnorm(n * p), n, p))

# Generate Poisson counts
lambda <- exp(X %*% true_beta)
y_pois <- rpois(n, lambda)
```

## Defining the Log-Likelihood

```{r}
poisson_loglik <- function(pars, y, X) {
  eta <- X %*% pars
  lambda <- exp(eta)
  sum(dpois(y, lambda, log = TRUE))
}
```

## Performance Comparison

```{r, results=FALSE, message=FALSE, warning=FALSE}
inits_pois <- rep(0, 3)

# Finite differences
timing_pois_fd <- system.time({
  fit_pois_fd <- stan_sample(poisson_loglik, inits_pois,
                             additional_args = list(y = y_pois, X = X),
                             num_chains = 1, seed = 1234)
})

# RTMB
timing_pois_rtmb <- system.time({
  fit_pois_rtmb <- stan_sample(poisson_loglik, inits_pois,
                               grad_fun = "RTMB",
                               additional_args = list(y = y_pois, X = X),
                               num_chains = 1, seed = 1234)
})
```

## Results

```{r}
timing_results_pois <- data.frame(
  Method = c("Finite Differences", "RTMB"),
  Time_seconds = c(timing_pois_fd[3], timing_pois_rtmb[3]),
  Speedup = c(1, timing_pois_fd[3] / timing_pois_rtmb[3])
)
knitr::kable(timing_results_pois, digits = 2,
             caption = "Performance comparison for Poisson regression")
```

```{r}
summary(fit_pois_rtmb)
```

RTMB handles the matrix operations and log-link function automatically, providing improved performance (typically 8-10x speedup) while correctly recovering the true parameter values.

# Logistic Regression

Logistic regression models binary outcomes using a logit link: $\text{logit}(p_i) = X_i\beta$, where $p_i$ is the probability of success.

## Simulating Data

```{r}
set.seed(789)
n <- 300
p <- 2

# True coefficients
true_beta_logit <- c(0.2, 1.5, -1.0)

# Design matrix
X_logit <- cbind(1, matrix(rnorm(n * p), n, p))

# Generate binary outcomes
eta <- X_logit %*% true_beta_logit
prob <- plogis(eta)  # inverse logit
y_binom <- rbinom(n, size = 1, prob = prob)
```

## Defining the Log-Likelihood

```{r}
logistic_loglik <- function(pars, y, X) {
  eta <- X %*% pars
  p <- plogis(eta)
  sum(dbinom(y, size = 1, prob = p, log = TRUE))
}
```

## Performance Comparison

```{r, results=FALSE, message=FALSE, warning=FALSE}
inits_logit <- rep(0, 3)

# Finite differences
timing_logit_fd <- system.time({
  fit_logit_fd <- stan_sample(logistic_loglik, inits_logit,
                              additional_args = list(y = y_binom, X = X_logit),
                              num_chains = 1, seed = 1234)
})

# RTMB
timing_logit_rtmb <- system.time({
  fit_logit_rtmb <- stan_sample(logistic_loglik, inits_logit,
                                grad_fun = "RTMB",
                                additional_args = list(y = y_binom, X = X_logit),
                                num_chains = 1, seed = 1234)
})
```

## Results

```{r}
timing_results_logit <- data.frame(
  Method = c("Finite Differences", "RTMB"),
  Time_seconds = c(timing_logit_fd[3], timing_logit_rtmb[3]),
  Speedup = c(1, timing_logit_fd[3] / timing_logit_rtmb[3])
)
knitr::kable(timing_results_logit, digits = 2,
             caption = "Performance comparison for Logistic regression")
```

```{r}
summary(fit_logit_rtmb)
```


# Gaussian Mixture Model

Mixture models represent complex latent structure and demonstrate RTMB's benefits for challenging models. We'll fit a two-component Gaussian mixture.

The model is: $p(y) = \pi \cdot N(\mu_1, \sigma_1^2) + (1-\pi) \cdot N(\mu_2, \sigma_2^2)$

## Simulating Data

```{r}
set.seed(101)
n <- 400

# True parameters
true_pi <- 0.3
true_mu1 <- -2
true_mu2 <- 3
true_sigma1 <- 1
true_sigma2 <- 1.5

# Generate mixture data
component <- rbinom(n, 1, true_pi)
y_mix <- ifelse(component == 1,
                rnorm(n, true_mu1, true_sigma1),
                rnorm(n, true_mu2, true_sigma2))
```

## Defining the Log-Likelihood

```{r}
mixture_loglik <- function(pars, y) {
  # Transform parameters to satisfy constraints
  pi <- pars[1]  # mixing proportion in [0,1]
  mu1 <- pars[2]
  mu2 <- pars[3]
  sigma1 <- pars[4]  # positive
  sigma2 <- pars[5]  # positive

  # Log-likelihood for each component
  log_lik1 <- dnorm(y, mu1, sigma1, log = TRUE) + log(pi)
  log_lik2 <- dnorm(y, mu2, sigma2, log = TRUE) + log(1 - pi)

  sum(log(exp(log_lik1) + exp(log_lik2)))
}
```

## Performance Comparison

```{r, results=FALSE, message=FALSE, warning=FALSE}
# Initialize near true values (mixture models can have multimodality)
inits_mix <- c(0.3, -2, 3, 1, 1.5)

# Finite differences
timing_mix_fd <- system.time({
  fit_mix_fd <- stan_sample(mixture_loglik, inits_mix,
                            lower = c(0, -Inf, -Inf, 0, 0),
                            upper = c(1, Inf, Inf, Inf, Inf),
                            additional_args = list(y = y_mix),
                            num_chains = 1, seed = 1234)
})

# RTMB
timing_mix_rtmb <- system.time({
  fit_mix_rtmb <- stan_sample(mixture_loglik, inits_mix,
                              lower = c(0, -Inf, -Inf, 0, 0),
                              upper = c(1, Inf, Inf, Inf, Inf),
                              grad_fun = "RTMB",
                              additional_args = list(y = y_mix),
                              num_chains = 1, seed = 1234)
})
```

## Results

```{r}
timing_results_mix <- data.frame(
  Method = c("Finite Differences", "RTMB"),
  Time_seconds = c(timing_mix_fd[3], timing_mix_rtmb[3]),
  Speedup = c(1, timing_mix_fd[3] / timing_mix_rtmb[3])
)
knitr::kable(timing_results_mix, digits = 2,
             caption = "Performance comparison for Gaussian Mixture")
```

```{r}
summary(fit_mix_rtmb)
```


# Time Series: AR(1) Model

An autoregressive model of order 1 (AR(1)) captures temporal dependence: $y_t = \phi y_{t-1} + \epsilon_t$, where $|\phi| < 1$ for stationarity.

## Simulating Data

```{r}
set.seed(202)
n <- 200
true_phi <- 0.7
true_sigma <- 1

# Simulate AR(1) process
y_ar <- numeric(n)
y_ar[1] <- rnorm(1, 0, true_sigma / sqrt(1 - true_phi^2))
for (t in 2:n) {
  y_ar[t] <- true_phi * y_ar[t-1] + rnorm(1, 0, true_sigma)
}
```

## Defining the Log-Likelihood

We use `tanh()` to constrain φ to (-1, 1).

```{r}
ar1_loglik <- function(pars, y) {
  phi <- pars[1]  # constrain to (-1, 1)
  sigma <-pars[2]  # positive

  n <- length(y)

  # First observation from stationary distribution
  ll <- dnorm(y[1], 0, sigma / sqrt(1 - phi^2), log = TRUE)

  # Subsequent observations
  for (t in 2:n) {
    ll <- ll + dnorm(y[t], phi * y[t-1], sigma, log = TRUE)
  }

  ll
}
```

## Performance Comparison

```{r, results=FALSE, message=FALSE, warning=FALSE}
inits_ar <- c(0.5, 1)

# Finite differences
timing_ar_fd <- system.time({
  fit_ar_fd <- stan_sample(ar1_loglik, inits_ar,
                           lower = c(-1, 0),
                           upper = c(0, Inf),
                           additional_args = list(y = y_ar),
                           num_chains = 1, seed = 1234)
})

# RTMB
timing_ar_rtmb <- system.time({
  fit_ar_rtmb <- stan_sample(ar1_loglik, inits_ar,
                             lower = c(-1, 0),
                             upper = c(0, Inf),
                             grad_fun = "RTMB",
                             additional_args = list(y = y_ar),
                             num_chains = 1, seed = 1234)
})
```

## Results

```{r}
timing_results_ar <- data.frame(
  Method = c("Finite Differences", "RTMB"),
  Time_seconds = c(timing_ar_fd[3], timing_ar_rtmb[3]),
  Speedup = c(1, timing_ar_fd[3] / timing_ar_rtmb[3])
)
knitr::kable(timing_results_ar, digits = 2,
             caption = "Performance comparison for AR(1) model")
```

```{r}
summary(fit_ar_rtmb)
```


## Quick Approximation with Pathfinder

RTMB also works with Pathfinder, Stan's fast variational inference method:

```{r, results=FALSE, message=FALSE, warning=FALSE}
fit_ar_path <- stan_pathfinder(ar1_loglik, inits_ar,
                               grad_fun = "RTMB",
                               additional_args = list(y = y_ar))
```

```{r}
summary(fit_ar_path)
```

